---
title: "LC analyses: subclustering"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Computational analysis pipeline for LC project: subclustering within TH expressing spots


# Analysis pipeline

## Load preprocessed data

Load preprocessed `SpatialExperiment` objects saved with previous script.

Note we have one `SpatialExperiment` object per sample, stored in a list.

```{r, message=FALSE}
library(SpatialExperiment)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
```

```{r}
# load objects
file_load <- here("outputs/objects/LC_spe_TH.RData")
load(file_load)
```

```{r}
# check
sapply(spe_TH, dim)
head(colData(spe_TH[[1]]))
```


# Check spots

Check spots have been identified in correct regions.

```{r, fig.width=3.5, fig.height=3.25}
# overall plots
sample_names <- names(spe_TH)

for (i in seq_along(sample_names)) {
  
  spe <- spe_TH[[i]]
  
  p <- ggplot(as.data.frame(cbind(colData(spe), spatialCoords(spe))), 
              aes(x = y, y = x, color = high_TH)) + 
    geom_point(size = 0.35) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_manual(values = c("gray90", "darkgreen")) + 
    ggtitle(paste0("Expressed TH: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


# Clustering analyses

Apply clustering pipeline within TH expressing spots only.

Using code from OSTA book.

Skip sample `Br5459_LC` since this one does not have enough TH expressing spots.

```{r}
# subset objects for simplicity
spe_sub <- vector("list", length = length(sample_names))
names(spe_sub) <- sample_names

for (i in seq_along(sample_names)) {
  spe <- spe_TH[[i]]
  ix_keep <- colData(spe)$high_TH
  spe_sub[[i]] <- spe[, ix_keep]
}

sapply(spe_sub, dim)
```


```{r}
#for (i in c(1:3, 5)) {
for (i in c(1, 5)) {
  
  spe <- spe_sub[[i]]
  
  # NORMALIZATION
  
  library(scran)
  # quick clustering for pool-based size factors
  set.seed(123)
  #qclus <- quickCluster(spe)
  # calculate size factors
  #spe <- computeSumFactors(spe, cluster = qclus)
  # calculate logcounts (log-transformed normalized counts)
  spe <- logNormCounts(spe)
  
  # FEATURE SELECTION
  
  # remove mitochondrial genes
  #spe <- spe[!is_mito, ]
  # fit mean-variance relationship
  dec <- modelGeneVar(spe)
  # select top HVGs
  top_hvgs <- getTopHVGs(dec, prop = 0.1)
  
  # DIMENSIONALITY REDUCTION
  
  library(scater)
  # compute PCA
  set.seed(123)
  spe <- runPCA(spe, subset_row = top_hvgs)
  # compute UMAP on top 50 PCs
  set.seed(123)
  spe <- runUMAP(spe, dimred = "PCA")
  # update column names
  colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
  
  # CLUSTERING
  
  # graph-based clustering
  set.seed(123)
  k <- 10
  g <- buildSNNGraph(spe, k = k, use.dimred = "PCA")
  g_walk <- igraph::cluster_walktrap(g)
  clus <- g_walk$membership
  colLabels(spe) <- factor(clus)
  
  # store objects
  spe_sub[[i]] <- spe
}
```


## Plots: UMAP

```{r, fig.width=3.5, fig.height=3.25}
# color palette
colors <- unname(palette.colors(palette = "Dark2"))

for (i in c(1, 5)) {
  
  spe <- spe_sub[[i]]
  
  p <- ggplot(as.data.frame(cbind(colData(spe), reducedDim(spe, type = "UMAP"))), 
              aes(x = UMAP1, y = UMAP2, color = label)) + 
    geom_point(size = 0.75) + 
    scale_color_manual(values = colors) + 
    ggtitle(paste0("Subclustering TH spots: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


## Plots: spatial

```{r, fig.width=3.5, fig.height=3.25}
for (i in c(1, 5)) {
  
  spe_sub_i <- spe_sub[[i]]
  spe <- spe_TH[[i]]
  colData(spe)[rownames(colData(spe_sub_i)), "label"] <- colData(spe_sub_i)$label
  n_clus <- length(table(colData(spe)$label))
  colData(spe)$label[is.na(colData(spe)$label)] <- "NA"
  colData(spe)$label <- as.factor(colData(spe)$label)
  
  p <- ggplot(as.data.frame(cbind(colData(spe), spatialCoords(spe))), 
              aes(x = y, y = x, color = label)) + 
    geom_point(size = 0.35) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_manual(values = c(colors[1:n_clus], "gray90")) + 
    ggtitle(paste0("Subclustering TH spots: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


# Session info

```{r}
sessionInfo()
```

