---
title: "LC analysis pipeline: spot-level cell type deconvolution"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Computational analysis pipeline for LC project: spot-level cell type deconvolution.

This script loads preprocessed `SpatialExperiment` objects from `LC_preprocessing.Rmd` and continues with downstream analyses on spot-level cell type deconvolution.

Samples: LC_1 and LC_2 (combined NextSeq and MiSeq reads)


# Analysis pipeline


## Load preprocessed data

Load preprocessed `SpatialExperiment` objects saved with `LC_preprocessing.Rmd` script.

Note we have one `SpatialExperiment` object per sample, stored in a list.

```{r, message=FALSE}
library(SpatialExperiment)
library(dplyr)
library(tidyr)
library(here)
```

```{r}
# load objects
file_load <- here("outputs/objects/LC_SPE_list.RData")
load(file_load)
```


## Load reference cell types

Load reference cell type marker gene expression profiles, provided by Louise Huuki.

This is a pan-brain set of cell types and marker gene expression profiles, consisting of 6 major cell types: astrocytes, microglia, oligodendrocytes, oligodendrocyte progenitor cells (OPCs), excitatory neurons, and inhibitory neurons (interneurons).

Marker genes for each cells are identified by ranking genes in terms of the relative difference in expression compared to the cell type with the next-highest expression of that gene. This strategy tends to give more specific markers than comparing against all other cells.

```{r}
# load marker gene set
file_markers <- here("inputs/marker_genes/marker_stats.Rdata")
load(file_markers)
```

```{r}
# use 'broad' set (i.e. major cell types)
head(marker_stats$broad)
# 'cellType.target' is the cell type of interest
table(marker_stats$broad$cellType.target)

# extract top 10 marker genes for each cell type
markers_top <- 
  marker_stats$broad %>% 
  filter(rank_ratio <= 10) %>% 
  select(c(gene, cellType.target, mean.target, rank_ratio, Symbol))

head(markers_top)
dim(markers_top)

# all markers are unique per cell type
length(unique(markers_top$Symbol))

# summary table
markers_top %>% 
  select(c("cellType.target", "rank_ratio", "Symbol")) %>% 
  spread(key = "cellType.target", value = "Symbol")
```


# Deconvolution methods

The following spot-level cell type deconvolution methods have recently been published. However, these are all difficult to apply in practice.

- RCTD: [Cable et al. (2021)](https://www.nature.com/articles/s41587-021-00830-w)

    - Requires matched snRNA-seq and ST data from the same samples, where the snRNA-seq data must consist of a full UMI count table. We do not have this here, so we cannot apply this method.

- SpatialDWLS: [Dong et al. (2021)](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02362-7)

    - Complicated to install, since it requires installing the full Giotto pipeline. Available as function `runDWLSDeconv` in Giotto package. No vignette or examples available.

- DestVI: [Lopez et al. (2021)](https://www.biorxiv.org/content/10.1101/2021.05.10.443517v1)

    - Available as a module in scVI in Python. Hard to integrate into R pipelines.

- STdeconvolve: Work in progress by [Brendan Miller and Jean Fan](https://github.com/JEFworks-Lab/STdeconvolve)

    - Based on LDA. Work in progress.

- SPOTlight: [Elosua-Bayes et al. (2021)](https://academic.oup.com/nar/article/49/9/e50/6129341)

    - Requires full scRNA-seq dataset as input and performs automated clustering and marker gene selection using Seurat.

List additional methods here.


# Non-negative matrix factorization (NMF)

Try our own unsupervised method using non-negative matrix factorization (NMF) (suggestion from Casey Greene).

Using `NMF` package from CRAN ([Gaujoux et al. 2010](https://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-11-367)).

```{r, message=FALSE}
library(NMF)
library(ggplot2)
```


## Additional preprocessing for NMF

Subset spots containing relatively high UMIs for further analysis.

Using sample-specific thresholds on UMIs per spot.

```{r, fig.width=4.75, fig.height=3.5}
# sample names
sample_names <- c("LC_1", "LC_2")

# select thresholds for each sample
thresh_umi <- c(3000, 1500)

for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  # identify spots above threshold
  colData(spe)$is_high_umi <- colData(spe)$sum > thresh_umi[i]
  # note: remove duplicated columns introduced when adding columns to colData
  colData(spe) <- colData(spe)[, !(duplicated(colnames(colData(spe))) | duplicated(colnames(colData(spe)), fromLast = TRUE))]
  spe_list[[i]] <- spe
  
  # plot thresholded regions
  p <- ggplot(as.data.frame(cbind(colData(spe), spatialCoords(spe))), 
              aes(x = y, y = x, color = is_high_umi)) + 
    geom_point(size = 0.5) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_manual(values = c("gray90", "red")) + 
    ggtitle(paste0("High UMI counts: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


Store subsetted objects in new list.

```{r}
# lists to store subsetted objects
spe_high_list <- spe_low_list <- vector("list", length = length(sample_names))
names(spe_high_list) <- names(spe_low_list) <- sample_names

for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  spe_high <- spe[, colData(spe)$is_high_umi]
  spe_low <- spe[, !colData(spe)$is_high_umi]
  
  dim(spe_high)
  dim(spe_low)
  
  spe_high_list[[i]] <- spe_high
  spe_low_list[[i]] <- spe_low
}
```


Normalization for spots within segmented region.

```{r, message=FALSE}
library(scater)
library(scran)
```

```{r, fig.width=4.5, fig.height=4}
for (i in seq_along(sample_names)) {
  
  # using subsetted object containing high UMIs
  spe <- spe_high_list[[i]]
  dim(spe)
  
  # calculate pool-based size factors
  set.seed(123)
  qclus <- quickCluster(spe)
  table(qclus)
  spe <- computeSumFactors(spe, cluster = qclus)
  
  summary(sizeFactors(spe))
  hist(log10(sizeFactors(spe)), 
       xlab = "log10 (size factors)", main = "Size factors")
  
  # calculate logcounts
  spe <- logNormCounts(spe)
  
  # store in list
  spe_high_list[[i]] <- spe
}
```


Feature selection to identify set of top highly variable genes (HVGs) for spots within segmented region.

```{r, fig.width=4.5, fig.height=4}
top_hvgs_list <- vector("list", length = length(sample_names))
names(top_hvgs_list) <- sample_names

for (i in seq_along(sample_names)) {
  
  # using subsetted object containing high UMIs
  spe <- spe_high_list[[i]]
  dim(spe)
  
  # remove mitochondrial genes
  is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
  table(is_mito)
  rowData(spe)$gene_name[is_mito]
  spe <- spe[!is_mito, ]
  dim(spe)
  
  # remove low-count genes
  is_low <- rowSums(counts(spe)) <= 5
  table(is_low)
  spe <- spe[!is_low, ]
  dim(spe)
  
  # store in list
  spe_high_list[[i]] <- spe
  
  # fit mean-variance relationship
  dec <- modelGeneVar(spe)
  fit <- metadata(dec)
  plot(fit$mean, fit$var, 
       xlab = "mean of log-expression", ylab = "variance of log-expression")
  curve(fit$trend(x), col = "dodgerblue", add = TRUE, lwd = 2)
  
  # select top HVGs
  top_hvgs <- getTopHVGs(dec, prop = 0.1)
  length(top_hvgs)
  
  # store in list
  top_hvgs_list[[i]] <- top_hvgs
  
  # store in SpatialExperiment object
  rowData(spe)$is_hvg <- rowData(spe)$gene_id %in% top_hvgs
  spe_high_list[[i]] <- spe
}
```


## Run NMF

Run NMF using set of top highly variable genes (HVGs) for spots within segmented region.

Runtime: seconds

```{r}
nmf_out_list <- vector("list", length = length(sample_names))
names(nmf_out_list) <- sample_names

for (i in seq_along(sample_names)) {
  
  # using subsetted object containing high UMIs
  spe <- spe_high_list[[i]]
  dim(spe)
  
  # keep rows containing HVGs
  spe_hvgs <- spe[rowData(spe)$is_hvg, ]
  dim(spe_hvgs)
  
  # run NMF using logcounts
  x <- as.matrix(logcounts(spe_hvgs))
  
  # assumed number of cell types
  r <- 10
  
  # run NMF using default algorithm and parameters (runtime: ~15 sec)
  nmf_out <- nmf(x, rank = r)
  
  # store in list
  nmf_out_list[[i]] <- nmf_out
  
  # check outputs
  # expression matrix (genes * NMF cell types)
  W <- basis(nmf_out)
  dim(W)
  head(W)
  # composition matrix (NMF cell types * spots)
  H <- coef(nmf_out)
  dim(H)
  H[, 1:6]
}
```


## Plot composition

Plot composition of NMF cell types per spot.

```{r, fig.width=3.5, fig.height=3}
for (i in seq_along(sample_names)) {
  
  print(sample_names[i])
  
  spe <- spe_high_list[[i]]
  nmf_out <- nmf_out_list[[i]]
  
  # composition matrix (transposed: spots * NMF cell types)
  H <- coef(nmf_out)
  df_H <- t(H)
  colnames(df_H) <- paste0("NMF_celltype_", seq_len(ncol(df_H)))
  head(df_H)
  stopifnot(all(rownames(df_H) == rownames(colData(spe))))
  
  # plot composition: one panel per NMF cell type
  for (k in seq_len(ncol(df_H))) {
    p <- ggplot(as.data.frame(cbind(colData(spe), spatialCoords(spe), prop = df_H[, k])), 
                aes(x = y, y = x, color = prop)) + 
      geom_point(size = 0.5) + 
      coord_fixed() + 
      scale_y_reverse() + 
      scale_color_gradient(low = "gray85", high = "blue") + 
      ggtitle(paste0(sample_names[i], ": ", colnames(df_H)[k])) + 
      guides(color = guide_colorbar("proportion\n(logcounts)")) + 
      theme_bw() + 
      theme(panel.grid = element_blank(), 
            axis.title = element_blank(), 
            axis.text = element_blank(), 
            axis.ticks = element_blank())
    print(p)
  }
}
```


## Plot expression profiles

Plot expression profiles of NMF cell types.

```{r, message=FALSE}
library(ComplexHeatmap)
```

```{r, fig.width=6, fig.height=12}
# threshold to display genes in heatmap
thresh_sum_logcounts <- c(700, 300)

for (i in seq_along(sample_names)) {
  
  print(sample_names[i])
  
  spe <- spe_high_list[[i]]
  # keep rows containing HVGs
  spe_hvgs <- spe[rowData(spe)$is_hvg, ]
  
  nmf_out <- nmf_out_list[[i]]
  
  # expression matrix (genes * NMF cell types)
  df_W <- basis(nmf_out)
  colnames(df_W) <- paste0("NMF_celltype_", seq_len(ncol(df_W)))
  head(df_W)
  stopifnot(all(rownames(df_W) == rownames(rowData(spe_hvgs))))
  rownames(df_W) <- rowData(spe_hvgs)$gene_name
  
  # keep only highly-expressed genes
  rowData(spe_hvgs)$sum_logcounts <- rowSums(logcounts(spe_hvgs))
  rowData(spe_hvgs)$is_high <- rowData(spe_hvgs)$sum_logcounts > thresh_sum_logcounts[i]
  df_W_high <- df_W[rowData(spe_hvgs)$is_high, ]
  dim(df_W_high)
  
  # plot heatmap of expression profiles
  print(Heatmap(df_W_high, name = sample_names[i]))
}
```


## Compare with known markers

Compare NMF cell type expression profiles with known marker gene sets.

Expression heatmap of known marker genes.

Note: In the current version (6 major cell types), we have likely excluded highly specific overall neuron markers, since markers are only ranked highly if they can distinguish excitatory from inhibitory neurons. To do: updated version using 5 major cell types instead (with combined neurons category).

```{r, fig.width=5, fig.height=8}
# expression matrix of known marker genes
df_markers_top <- markers_top[, c("Symbol", "cellType.target", "mean.target")]
colnames(df_markers_top) <- c("gene_name", "celltype", "mean")

df_markers_top <- as.data.frame(spread(df_markers_top, "celltype", "mean"))

rownames(df_markers_top) <- df_markers_top$gene_name
df_markers_top <- df_markers_top[, -1]
head(df_markers_top)

# set NA values to 0 for heatmaps
mat_markers_top <- as.matrix(df_markers_top)
mat_markers_top[is.na(mat_markers_top)] <- 0

Heatmap(mat_markers_top, 
        cluster_columns = FALSE, name = "markers", 
        row_names_gp = gpar(fontsize = 10))
```


What is the overlap between the set of top HVGs (used for NMF) and known marker genes?

```{r}
for (i in seq_along(sample_names)) {
  
  print(sample_names[i])
  
  # set of top HVGs
  top_hvgs <- top_hvgs_list[[i]]
  
  # number of marker genes
  print(length(markers_top$gene))
  # number of HVGs for this sample
  print(length(top_hvgs))
  # number overlapping
  print(sum(markers_top$gene %in% top_hvgs))
}
```


The overlap between the set of top HVGs and the set of marker genes is very low, which may explain why the NMF cell types do not match well to the known cell types.


For comparison: What is the overlap between the set of all detected genes and known marker genes?

```{r}
for (i in seq_along(sample_names)) {
  
  print(sample_names[i])
  
  # using subsetted object containing high UMIs
  spe <- spe_high_list[[i]]
  
  # number of marker genes
  print(length(markers_top$gene))
  # number of detected genes for this sample
  print(nrow(spe))
  # number overlapping
  print(sum(markers_top$gene %in% rowData(spe)$gene_id))
}
```



# Non-negative least squares (NNLS)

Try our own supervised method using non-negative least squares (NNLS).

Using `nnls` package from CRAN ([Mullen et al. 2012](https://cran.r-project.org/web/packages/nnls/index.html)).

```{r, message=FALSE}
library(nnls)
```


## Run NNLS

Run NNLS in a loop (once per spot within segmented region) using the set of known marker genes.

Using (i) binary matrix of marker genes expressed per cell type sourced from external snRNA-seq data, then (ii) scale expression values per marker gene by mean log-expression per spot where this gene is detected in the ST data (so the reference marker expression profiles are on the correct scale for the ST data - suggestion from Shijie Zheng).

Note we detect only some of the marker genes per sample (see previous section).

Runtime: seconds

```{r}
# set of marker gene IDs (using gene ID instead of symbol for dis-ambiguity)
marker_ids_all <- markers_top$gene
length(marker_ids_all)

# create binary matrix of marker genes expressed per cell type
df_markers_top <- markers_top[, c("gene", "cellType.target", "mean.target")]
colnames(df_markers_top) <- c("gene_id", "celltype", "mean")
df_markers_top <- as.data.frame(spread(df_markers_top, "celltype", "mean"))
rownames(df_markers_top) <- df_markers_top$gene_id
df_markers_top <- df_markers_top[, -1]
df_markers_top[is.na(df_markers_top)] <- 0

markers_top_binary <- as.matrix(df_markers_top > 0) + 0
dim(markers_top_binary)
head(markers_top_binary)
```

```{r}
# run NNLS

nnls_out_list <- vector("list", length = length(sample_names))
names(nnls_out_list) <- sample_names

for (i in seq_along(sample_names)) {
  
  # using subsetted object containing high UMIs
  spe <- spe_high_list[[i]]
  print(dim(spe))
  
  # scale binary reference profile matrix by mean expression in spots where each 
  # gene is detected in ST data (so reference profiles are on same scale as ST data)
  
  # subset detected marker genes
  marker_ids <- marker_ids_all[marker_ids_all %in% rownames(spe)]
  length(marker_ids)
  markers_top_binary_sub <- markers_top_binary[marker_ids, ]
  print(dim(markers_top_binary_sub))
  
  # subset SPE object to keep only marker genes
  spe_markers <- spe[rowData(spe)$gene_id %in% marker_ids, ]
  print(dim(spe_markers))
  
  # scale reference matrix by mean log-expression per gene
  # note: treat non-expressed as NAs
  logcounts_nas <- as.matrix(logcounts(spe_markers))
  logcounts_nas[logcounts_nas == 0] <- NA
  mean_markers <- rowMeans(logcounts_nas, na.rm = TRUE)
  
  markers_scaled <- markers_top_binary_sub * 
                    matrix(rep(mean_markers, ncol(markers_top_binary_sub)), 
                           ncol = ncol(markers_top_binary_sub))
  head(markers_scaled)
  
  stopifnot(all(rownames(spe_markers) == rownames(spe_markers)))
  
  
  # loop over spots
  
  n_celltypes <- ncol(markers_scaled)
  
  nnls_out_mat <- matrix(NA, nrow = n_celltypes, ncol = ncol(spe_markers))
  rownames(nnls_out_mat) <- colnames(markers_scaled)
  colnames(nnls_out_mat) <- colnames(spe_markers)
  
  for (s in seq_len(ncol(spe_markers))) {
    # NNLS formula: A*x = b
    A <- markers_scaled
    b <- logcounts(spe_markers)[, s]
    # run NNLS
    x <- nnls(A, b)$x
    # store cell type proportions
    nnls_out_mat[, s] <- x
  }
  
  # store output
  nnls_out_list[[i]] <- nnls_out_mat
}
```


## Cell type proportions

Should cell type proportions per spot sum to one? Can we simply normalize them?

```{r, fig.width=5, fig.height=4}
for (i in seq_along(sample_names)) {
  
  print(sample_names[i])
  
  # histogram of cell type proportions per spot
  print(hist(colSums(nnls_out_list[[i]]), 
             main = sample_names[i], 
             xlab = "sum of cell type proportions per spot"))
}
```


## Plot cell type proportions

Plot cell type proportions per spot.

```{r, fig.width=3.5, fig.height=3}
for (i in seq_along(sample_names)) {
  
  print(sample_names[i])
  
  spe <- spe_high_list[[i]]
  nnls_out_mat <- nnls_out_list[[i]]
  
  # plot composition: one panel per cell type
  for (k in seq_len(nrow(nnls_out_mat))) {
    p <- ggplot(as.data.frame(cbind(colData(spe), spatialCoords(spe), prop = nnls_out_mat[k, ])), 
                aes(x = y, y = x, color = prop)) + 
      geom_point(size = 0.5) + 
      coord_fixed() + 
      scale_y_reverse() + 
      scale_color_gradient(low = "gray85", high = "blue") + 
      ggtitle(paste0(sample_names[i], ": ", rownames(nnls_out_mat)[k])) + 
      guides(color = guide_colorbar("proportion\n(logcounts)")) + 
      theme_bw() + 
      theme(panel.grid = element_blank(), 
            axis.title = element_blank(), 
            axis.text = element_blank(), 
            axis.ticks = element_blank())
    print(p)
  }
}
```


## Session info

```{r}
sessionInfo()
```

