---
title: "LC analyses: mouse markers"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Computational analysis pipeline for LC project: mouse marker genes.

Investigate expression of top highly expressed genes in mouse LC described by Mulvey et al. (2018) (converted to approximate human genes).


# Analysis pipeline


## Load preprocessed data

Load preprocessed `SpatialExperiment` objects saved with `LC_preprocessing_summaries.Rmd` script.

Note we have one `SpatialExperiment` object per sample, stored in a list.

```{r, message=FALSE}
library(SpatialExperiment)
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
```

```{r}
# load objects
file_load <- here("outputs/objects/LC_SPE_list.RData")
load(file_load)
```


## Load mouse marker genes

Using panel of top 45 marker genes for mouse LC, from Mulvey et al. (2018), Figure 2A: https://pubmed.ncbi.nlm.nih.gov/29791834/.

I have extracted the gene names from Figure 2A and saved them in a text file.

```{r, message=FALSE}
# load list of mouse marker genes
file_markers <- file.path(here("inputs/marker_genes/Mulvey2018_Fig2A_marker_genes.txt"))
mouse_markers <- read.table(file_markers)[, 1]

length(mouse_markers)
mouse_markers
```


## Convert to human genes

Convert list of mouse marker genes to homologous human gene names.

Using code from `biomaRt` vignette and https://www.r-bloggers.com/2016/10/converting-mouse-to-human-gene-names-with-biomart-package/

This retrieves homologous human gene names for 41 out of the 45 mouse marker genes from the list above.

```{r, message=FALSE}
library(biomaRt)
```

```{r, message=FALSE}
human = useMart("ensembl", dataset = "hsapiens_gene_ensembl")
mouse = useMart("ensembl", dataset = "mmusculus_gene_ensembl")

genes_v2 <- getLDS(
  attributes = c("mgi_symbol"), filters = "mgi_symbol", values = mouse_markers, 
  mart = mouse, attributesL = c("hgnc_symbol"), martL = human, uniqueRows = TRUE
)

head(genes_v2)

human_markers <- genes_v2[, 2]
human_markers <- sort(human_markers)

length(human_markers)
human_markers
```


# Plot expression

Plot UMI counts for the 41 homologous human genes matching to the top 45 mouse LC marker genes from Mulvey et al. (2018).

Organized as one subsection per sample below.

```{r}
# retrieve sample names
sample_names <- names(spe_list)

# function for plots
f_plots <- function(i, human_markers) {
  
  print(sample_names[i])
  
  spe <- spe_list[[i]]
  
  for (g in seq_along(human_markers)) {
    
    # identify gene of interest
    gn <- human_markers[g]
    ix <- which(rowData(spe)$gene_name == gn)
    colData(spe)$sum_gn <- counts(spe)[ix, ]
    # note: remove duplicated columns introduced when adding columns to colData
    colData(spe) <- colData(spe)[, !(duplicated(colnames(colData(spe))) | duplicated(colnames(colData(spe)), fromLast = TRUE))]
    
    # filter out low-expressed genes (max UMI counts per spot <= 1)
    if (max(counts(spe)[ix, ]) <= 1) next
    
    p <- ggplot(as.data.frame(cbind(colData(spe), spatialCoords(spe))), 
                aes(x = y, y = x, color = sum_gn)) + 
      geom_point(size = 0.3) + 
      coord_fixed() + 
      scale_y_reverse() + 
      scale_color_gradient(low = "gray90", high = "blue") + 
      labs(color = "UMI\ncounts") + 
      ggtitle(paste0(sample_names[i], ": ", gn)) + 
      theme_bw() + 
      theme(panel.grid = element_blank(), 
            axis.title = element_blank(), 
            axis.text = element_blank(), 
            axis.ticks = element_blank())
    
    print(p)
  }
}
```


## Br6522_LC_1

```{r, fig.width=2.75, fig.height=2.25}
f_plots(1, human_markers)
```


## Br6522_LC_2

```{r, fig.width=2.75, fig.height=2.25}
f_plots(2, human_markers)
```


## Br8153_LC

```{r, fig.width=2.75, fig.height=2.25}
f_plots(3, human_markers)
```


## Br5459_LC

```{r, fig.width=2.75, fig.height=2.25}
f_plots(4, human_markers)
```


## Br2701_LC

```{r, fig.width=2.75, fig.height=2.25}
f_plots(5, human_markers)
```


# Session info

```{r}
sessionInfo()
```

