---
title: "LC analyses: preprocessing distributions"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Computational analysis pipeline for LC project: preprocessing summaries and distributions


# Analysis pipeline

## Load data

Load data from Space Ranger and create `SpatialExperiment` objects.

Note that we create a separate `SpatialExperiment` object for each sample, and store these in a list.

```{r, message=FALSE}
library(SpatialExperiment)
library(Matrix)
library(rjson)
library(dplyr)
library(here)
```

```{r}
sample_names <- c("Br6522_LC_1", "Br6522_LC_2", "Br8153_LC", "Br5459_LC", "Br2701_LC")
spe_list <- vector("list", length = length(sample_names))
names(spe_list) <- sample_names

for (i in seq_along(sample_names)) {
  
  # ---------
  # Load data
  # ---------
  
  sample_name <- sample_names[i]
  
  # path to Space Ranger output files (either on JHPCE or Mac laptop)
  if (Sys.info()["sysname"] == "Linux") {
    # files on JHPCE cluster
    dir_outputs <- here("processed_data/spaceranger")
  } else if (Sys.info()["sysname"] == "Darwin") {
    # files on Mac laptop
    dir_outputs <- "~/data/locus-c/processed_data/spaceranger"
  }
  if (i %in% 1:2) {
    dir_outputs <- file.path(dir_outputs, "NextSeqMiSeq")
  } else if (i %in% 3:5) {
    dir_outputs <- file.path(dir_outputs, "Linda_2021-05-21")
  }
  
  # using "filtered" output files containing only spots over tissue
  dir_matrix <- file.path(dir_outputs, sample_name, "outs", "filtered_feature_bc_matrix")
  
  # barcodes
  file_barcodes <- file.path(dir_matrix, "barcodes.tsv.gz")
  df_barcodes <- read.csv(file_barcodes, sep = "\t", header = FALSE, 
                          col.names = c("barcode_id"))
  # features
  file_features <- file.path(dir_matrix, "features.tsv.gz")
  df_features <- read.csv(file_features, sep = "\t", header = FALSE, 
                          col.names = c("gene_id", "gene_name", "feature_type"))
  # counts
  file_counts <- file.path(dir_matrix, "matrix.mtx.gz")
  counts <- readMM(file = file_counts)
  
  # spatial coordinates
  dir_spatial <- file.path(dir_outputs, sample_name, "outs", "spatial")
  file_tisspos <- file.path(dir_spatial, "tissue_positions_list.csv")
  df_tisspos <- read.csv(file_tisspos, header = FALSE, 
                         col.names=c("barcode_id", "in_tissue", "array_col", "array_row", 
                                     "pxl_col_in_fullres", "pxl_row_in_fullres"))
  
  # image file paths
  img_paths <- c(
    aligned_fiducials = file.path(dir_spatial, "aligned_fiducials.jpg"), 
    detected_tissue_image = file.path(dir_spatial, "detected_tissue_image.jpg"), 
    tissue_hires_image = file.path(dir_spatial, "tissue_hires_image.png"), 
    tissue_lowres_image = file.path(dir_spatial, "tissue_lowres_image.png")
  )
  
  # spatial scale factors
  file_scale_factors <- file.path(dir_spatial, "scalefactors_json.json")
  scale_factors <- fromJSON(file = file_scale_factors)
  
  # check dimensions
  dim(df_barcodes)
  dim(df_features)
  dim(counts)
  # df_tisspos contains all spots (not filtered) - need to match rows
  dim(df_tisspos)
  
  
  # --------------
  # Match barcodes
  # --------------
  
  # match barcode IDs in 'df_barcodes' and 'df_tisspos'
  dim(df_barcodes)
  dim(df_tisspos)
  stopifnot(all(df_barcodes$barcode_id %in% df_tisspos$barcode_id))
  
  df_tisspos_ord <- left_join(df_barcodes, df_tisspos, by = "barcode_id")
  rownames(df_tisspos_ord) <- df_tisspos_ord$barcode_id
  dim(df_tisspos_ord)
  head(df_tisspos_ord, 3)
  
  
  # ------------------------
  # Create SpatialExperiment
  # ------------------------
  
  # row data
  row_data <- DataFrame(df_features)
  rownames(row_data) <- df_features$gene_id
  head(row_data, 3)
  
  # column data
  col_data <- DataFrame(sample_id = rep(sample_names[i], nrow(df_barcodes)))
  rownames(col_data) <- df_barcodes$barcode_id
  # note: also include duplicate spatial coordinates with original column names
  stopifnot(all(rownames(col_data) == rownames(df_tisspos_ord)))
  col_data <- cbind(col_data, df_tisspos_ord[, c("pxl_col_in_fullres", "pxl_row_in_fullres")])
  head(col_data, 3)
  
  # spatial coordinates
  spatial_coords <- as.matrix(df_tisspos_ord[, c("pxl_col_in_fullres", "pxl_row_in_fullres")])
  colnames(spatial_coords) <- c("x", "y")
  head(spatial_coords, 3)
  
  # spatial data
  spatial_data <- DataFrame(df_tisspos_ord)[, 1:4]
  rownames(spatial_data) <- df_tisspos_ord$barcode_id
  head(spatial_data, 3)
  
  # image data: low and high resolution images from Space Ranger
  img_data <- readImgData(
    path = dir_spatial, 
    sample_id = sample_names[i], 
    imageSources = c(img_paths["tissue_lowres_image"], img_paths["tissue_hires_image"]), 
    scaleFactors = file_scale_factors, 
    load = TRUE
  )
  
  # create SpatialExperiment
  spe <- SpatialExperiment(
    assays = list(counts = counts), 
    rowData = row_data, 
    colData = col_data, 
    spatialData = spatial_data, 
    spatialCoords = spatial_coords, 
    imgData = img_data
  )
  
  # store in list
  spe_list[[i]] <- spe
}

length(spe_list)
names(spe_list)
sapply(spe_list, dim)
```


# Quality control (QC) metrics

Calculate quality control (QC) metrics per sample using `scater`.

```{r, message=FALSE}
library(scater)
library(ggplot2)
```

```{r}
for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  # identify mitochondrial genes
  is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
  table(is_mito)
  rowData(spe)$gene_name[is_mito]
  
  # remove any spots with zero reads
  is_zero <- colSums(counts(spe)) == 0
  if (sum(is_zero) > 0) {
    spe <- spe[, !is_zero]
  }
  
  # calculate QC metrics using scater package
  spe <- addPerCellQC(spe, subsets = list(mito = is_mito))
  
  # note: remove duplicated columns introduced when adding columns to colData
  colData(spe) <- colData(spe)[, !(duplicated(colnames(colData(spe))) | duplicated(colnames(colData(spe)), fromLast = TRUE))]
  head(colData(spe), 3)
  
  # store in list
  spe_list[[i]] <- spe
}

# check
head(colData(spe_list[[1]]), 3)

length(spe_list)
names(spe_list)
sapply(spe_list, dim)
```


## Total UMI counts

```{r, fig.width=3.75, fig.height=3.25}
# plot UMI counts
for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  p <- ggplot(as.data.frame(cbind(colData(spe), spatialCoords(spe))), 
              aes(x = y, y = x, color = sum)) + 
    geom_point(size = 0.5) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient(low = "gray90", high = "navy") + 
    labs(color = "sum_umi") + 
    ggtitle(paste0("UMI counts: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


## Number of detected genes

```{r, fig.width=3.75, fig.height=3.25}
for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  p <- ggplot(as.data.frame(cbind(colData(spe), spatialCoords(spe))), 
              aes(x = y, y = x, color = detected)) + 
    geom_point(size = 0.5) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient(low = "gray90", high = "deeppink3") + 
    labs(color = "detected") + 
    ggtitle(paste0("Detected genes: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


## Mitochondrial percentage

```{r, fig.width=3.75, fig.height=3.25}
# plot percent mitochondrial reads
for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  p <- ggplot(as.data.frame(cbind(colData(spe), spatialCoords(spe))), 
              aes(x = y, y = x, color = subsets_mito_percent)) + 
    geom_point(size = 0.5) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient(limits = c(0, 100), low = "gray90", high = "darkorange2") + 
    labs(color = "mito") + 
    ggtitle(paste0("Percent mitochondrial: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


## High mitochondrial spots

```{r, fig.width=3.75, fig.height=3.25}
# spots with high mitochondrial reads
for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  # threshold percent mitochondrial reads
  thresh_mito <- 35
  colData(spe)$high_mito <- colData(spe)$subsets_mito_percent > thresh_mito
  
  p <- ggplot(as.data.frame(cbind(colData(spe), spatialCoords(spe))), 
              aes(x = y, y = x, color = high_mito)) + 
    geom_point(size = 0.35) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_manual(values = c("gray90", "darkorange2")) + 
    labs(color = "high_mito") + 
    ggtitle(paste0("High mitochondrial percent: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


# Distributions of UMI counts

## Spots with high UMI counts

Select threshold for checking distributions in LC region vs. background. Selected conservatively, i.e. prefer to select too many spots than too few.

```{r, fig.width=3.5, fig.height=3.25}
# new list to store objects
spe_highUMI <- vector("list", length = length(sample_names))
names(spe_highUMI) <- sample_names

# select thresholds
thresh_umi <- c(1000, 1000, 1000, 1000, 1000)

for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  colData(spe)$high_umi <- colData(spe)$sum > thresh_umi[i]
  
  # store object
  spe_list[[i]] <- spe
  spe_highUMI[[i]] <- spe
  
  p <- ggplot(as.data.frame(cbind(colData(spe), spatialCoords(spe))), 
              aes(x = y, y = x, color = high_umi)) + 
    geom_point(size = 0.35) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_manual(values = c("gray90", "red")) + 
    labs(color = "high_umi") + 
    ggtitle(paste0("High UMIs: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


Distributions for selected spots with high UMIs.

```{r, fig.width=3.5, fig.height=3.25}
# distributions of UMI counts for selected spots
for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  spe_sub <- spe[, colData(spe)$high_umi]
  
  hist(log10(colData(spe_sub)$sum), breaks = 20, 
       xlab = "log10 sum UMIs", main = sample_names[i], 
       col = "navy")
}
```


## Spots with detected TH expression

```{r, fig.width=3.5, fig.height=3.25}
# new list to store objects
spe_TH <- vector("list", length = length(sample_names))
names(spe_TH) <- sample_names

# select thresholds
thresh_TH <- 1

for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  ix_TH <- which(rowData(spe)$gene_name == "TH")
  colData(spe)$high_TH <- counts(spe)[ix_TH, ] > thresh_TH
  
  # store object
  spe_list[[i]] <- spe
  spe_TH[[i]] <- spe
  
  p <- ggplot(as.data.frame(cbind(colData(spe), spatialCoords(spe))), 
              aes(x = y, y = x, color = high_TH)) + 
    geom_point(size = 0.35) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_manual(values = c("gray90", "darkgreen")) + 
    ggtitle(paste0("Expressed TH: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


Distributions for selected spots with detected expression of TH.

```{r, fig.width=3.5, fig.height=3.25}
# distributions of UMI counts for selected spots
for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  spe_sub <- spe[, colData(spe)$high_TH]
  
  hist(log10(colData(spe_sub)$sum), breaks = 20, 
       xlab = "log10 sum UMIs", main = sample_names[i], 
       col = "navy")
}
```


# Save objects

Save `SpatialExperiment` objects for use in downstream analyses in other `RMarkdown` files.

```{r}
# save objects
file_out <- here("outputs/objects/LC_spe_highUMI.RData")
save(spe_highUMI, file = file_out)

file_out <- here("outputs/objects/LC_spe_TH.RData")
save(spe_TH, file = file_out)
```


# Session info

```{r}
sessionInfo()
```

