---
title: "LC analyses: white matter"
author: "Lukas Weber"
date: "19 Nov 2020"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction

Analyses for LC project: white matter.

Investigate expression within white matter region in each sample.


## Data

Using combined NextSeq and MiSeq reads for the LC samples (LC_1 and LC_2).


# Analysis

## Load data

See previous file `2_features_per_spot_LC.Rmd` for more details.

```{r, message=FALSE}
library(here)
library(tidyverse)
library(ggplot2)
library(SingleCellExperiment)
library(Matrix)
library(rjson)
```

```{r}
sample_names <- c("LC_1", "LC_2")
sce_list <- vector("list", length = length(sample_names))
names(sce_list) <- sample_names

for (i in seq_along(sample_names)) {
  
  # ---------
  # load data
  # ---------
  
  # select sample
  sample_name <- sample_names[i]
  
  # path to Space Ranger output files
  if (Sys.info()["sysname"] == "Linux") {
    # files on JHPCE cluster
    dir_outputs <- "/dcl02/lieber/ajaffe/SpatialTranscriptomics/LIBD/locus-c/outputs/NextSeqMiSeq"
  } else if (Sys.info()["sysname"] == "Darwin") {
    # copy of files on Mac laptop
    dir_outputs <- "~/data/locus-c/outputs/NextSeqMiSeq"
  }
  
  # note: using "filtered" barcodes list containing only spots over tissue
  dir_matrix <- file.path(dir_outputs, sample_name, "outs", "filtered_feature_bc_matrix")
  
  # barcodes
  file_barcodes <- file.path(dir_matrix, "barcodes.tsv.gz")
  df_barcodes <- read.csv(file_barcodes, sep = "\t", header = FALSE, 
                          col.names = c("barcode_id"))
  # features
  file_features <- file.path(dir_matrix, "features.tsv.gz")
  df_features <- read.csv(file_features, sep = "\t", header = FALSE, 
                          col.names = c("gene_id", "gene_name", "feature_type"))
  # counts
  file_counts <- file.path(dir_matrix, "matrix.mtx.gz")
  counts <- readMM(file = file_counts)
  
  # spatial coordinates
  dir_spatial <- file.path(dir_outputs, sample_name, "outs", "spatial")
  file_tisspos <- file.path(dir_spatial, "tissue_positions_list.csv")
  df_tisspos <- read.csv(file_tisspos, header = FALSE, 
                         col.names=c("barcode_id", "in_tissue", "array_row", "array_col", 
                                     "pxl_col_in_fullres", "pxl_row_in_fullres"))
  
  # check dimensions
  dim(df_barcodes)
  dim(df_features)
  dim(counts)
  # note df_tisspos contains all spots (not filtered) - need to match later
  dim(df_tisspos)
  
  # image paths
  imageFilePath <- file.path(dir_spatial, c("tissue_hires_image.png", "tissue_lores_image.png"))
  
  # spatial scale factors
  file_scale <- file.path(dir_spatial, "scalefactors_json.json")
  scalefactors <- fromJSON(file = file_scale)
  
  
  # ---------------------------
  # create SingleCellExperiment
  # ---------------------------
  
  # note: check and/or re-order rows to make sure barcode IDs match in df_barcodes and df_tisspos
  dim(df_barcodes)
  dim(df_tisspos)
  ord <- match(df_barcodes$barcode_id, df_tisspos$barcode_id)
  df_tisspos_ord <- df_tisspos[ord, ]
  dim(df_tisspos_ord)
  stopifnot(nrow(df_barcodes) == nrow(df_tisspos_ord))
  stopifnot(all(df_barcodes$barcode_id == df_tisspos_ord$barcode_id))
  
  head(df_barcodes)
  head(df_tisspos_ord)
  
  col_data <- cbind(df_barcodes, df_tisspos_ord[, -1])
  head(col_data)
  
  sce <- SingleCellExperiment(
    rowData = df_features, 
    colData = col_data, 
    assays = c(counts = counts), 
    metadata = list(scalefactors = scalefactors)
  )
  
  sce
  
  # store object
  sce_list[[i]] <- sce
}

sce_list
```


## Calculate scater QC metrics

```{r, message=FALSE}
library(scater)
```

```{r, message=FALSE}
for (i in seq_along(sample_names)) {
  
  # select sample
  sce <- sce_list[[i]]
  
  # identify mitochondrial genes
  is_mito <- grepl("(^MT-)|(^mt-)", rowData(sce)$gene_name)
  table(is_mito)
  rowData(sce)$gene_name[is_mito]
  
  # calculate QC metrics using scater package
  sce <- addPerCellQC(sce, subsets = list(mito = is_mito))
  
  colData(sce)
  
  # store
  sce_list[[i]] <- sce
}
```


## Plot white matter marker genes

Plot UMI counts for white matter marker genes: MBP

```{r}
dir_plot <- here("plots", "white_matter")
if (!dir.exists(dir_plot)) dir.create(dir_plot, recursive = TRUE)
```

```{r}
# white matter marker genes
wm_markers <- c("MBP", "PCP4")
```

```{r, message=FALSE, fig.width=3.5, fig.height=2.5}
# plot UMI counts
colors <- c("navy", "dodgerblue2")

for (i in seq_along(sample_names)) {
  
  # select sample
  sce <- sce_list[[i]]
  
  for (j in seq_along(wm_markers)) {
    
    # identify marker gene
    ix_marker <- which(toupper(rowData(sce)$gene_name) == toupper(wm_markers[j]))
    stopifnot(length(ix_marker) == 1)
    colData(sce)$counts_marker <- counts(sce)[ix_marker, ]
    
    
    # plot UMI counts for marker gene
    
    p <- ggplot(as.data.frame(colData(sce)), 
                aes(x = pxl_row_in_fullres, y = pxl_col_in_fullres, color = counts_marker)) + 
      geom_point(size = 0.25) + 
      coord_fixed() + 
      scale_y_reverse() + 
      scale_color_gradient(low = "gray95", high = colors[j]) + 
      ggtitle(paste0("UMI counts: ", wm_markers[j], ": ", sample_names[i])) + 
      labs(color = "counts") + 
      theme_bw() + 
      theme(panel.grid = element_blank(), 
            axis.title = element_blank(), 
            axis.text = element_blank(), 
            axis.ticks = element_blank())
    
    print(p)
    
    fn <- file.path(dir_plot, paste0("xyplot_counts_", wm_markers[j], "_", sample_names[i], ".png"))
    ggsave(fn, width = 4.75, height = 3.5)
  }
}
```


## Compare with DLPFC samples

Load DLPFC samples from `spatialLIBD`.

```{r}
# load spatialLIBD object
library(spatialLIBD)
sce_dlpfc <- fetch_data(type = "sce")

# ground truth layer labels
table(sce_dlpfc$layer_guess_reordered, useNA = "always")
```

```{r, message=FALSE, fig.width=3.5, fig.height=2.5}
# spatialLIBD plots
sce_image_clus(
  sce = sce_dlpfc, 
  clustervar = "layer_guess_reordered", 
  sampleid = "151673", 
  colors = libd_layer_colors
)
```

```{r, message=FALSE, fig.width=3.5, fig.height=2.5}
# UMI counts for white matter

# identify white matter spots
ix_wm <- colData(sce_dlpfc)$layer_guess_reordered == "WM"
ix_wm[is.na(ix_wm)] <- FALSE

colData(sce_dlpfc)$ix_wm <- ix_wm

# object with white matter spots
sce_wm <- sce_dlpfc[, colData(sce_dlpfc)$ix_wm]
dim(sce_wm)

# object with non white matter spots
sce_nonwm <- sce_dlpfc[, !colData(sce_dlpfc)$ix_wm]
dim(sce_nonwm)


# histograms: sample 151673 only
fn <- file.path(dir_plot, paste0("hist_UMI_counts_per_spot_DLPFC_WM_151673.png"))
png(fn)
hist(colSums(counts(sce_wm[, colData(sce_wm)$sample_name == "151673"])), 
     main = "UMI counts for WM spots (sample 151673)", 
     xlab = "total UMI counts")
dev.off()

fn <- file.path(dir_plot, paste0("hist_UMI_counts_per_spot_DLPFC_nonWM_151673.png"))
png(fn)
hist(colSums(counts(sce_nonwm[, colData(sce_nonwm)$sample_name == "151673"])), 
     main = "UMI counts for non-WM spots (sample 151673)", 
     xlab = "total UMI counts")
dev.off()


# histograms: all samples
fn <- file.path(dir_plot, paste0("hist_UMI_counts_per_spot_DLPFC_WM_all.png"))
png(fn)
hist(colSums(counts(sce_wm)), 
     main = "UMI counts for WM spots (all samples)", 
     xlab = "total UMI counts")
dev.off()

fn <- file.path(dir_plot, paste0("hist_UMI_counts_per_spot_DLPFC_nonWM_all.png"))
png(fn)
hist(colSums(counts(sce_nonwm)), 
     main = "UMI counts for non-WM spots (all samples)", 
     xlab = "total UMI counts")
dev.off()
```

```{r, message=FALSE, fig.width=3.5, fig.height=2.5}
# plot total UMI counts (showing white matter region)
sce_sub <- sce_dlpfc[, colData(sce_dlpfc)$sample_name == "151673"]
dim(sce_sub)

p <- ggplot(as.data.frame(colData(sce_sub)), 
            aes(x = imagecol, y = -1 * imagerow, color = sum_umi)) + 
  geom_point(size = 0.5) + 
  coord_fixed() + 
  scale_color_gradient(low = "gray95", high = "navy") + 
  ggtitle(paste0("UMI counts per spot: sample 151673")) + 
  theme_bw() + 
  theme(panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank())

print(p)

fn <- file.path(dir_plot, paste0("xyplot_UMI_counts_per_spot_151673.png"))
ggsave(fn, width = 4.75, height = 3.5)


# plot number of detected genes (showing white matter region)
p <- ggplot(as.data.frame(colData(sce_sub)), 
            aes(x = imagecol, y = -1 * imagerow, color = sum_gene)) + 
  geom_point(size = 0.5) + 
  coord_fixed() + 
  scale_color_gradient(low = "gray95", high = "darkgreen") + 
  ggtitle(paste0("Detected genes per spot: sample 151673")) + 
  theme_bw() + 
  theme(panel.grid = element_blank(), 
        axis.title = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank())

print(p)

fn <- file.path(dir_plot, paste0("xyplot_detected genes_per_spot_151673.png"))
ggsave(fn, width = 4.75, height = 3.5)
```


# Session info

```{r, message=FALSE}
sessionInfo()
```

