---
title: "LC analysis pipeline: segmented regions"
author: "Lukas Weber"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


# Introduction

Analyses for LC project: analyses within segmented regions containing neurons and white matter.


## Data

Using combined NextSeq and MiSeq reads for LC samples LC_1 and LC_2.


# Analysis pipeline

## Load data

Load data from Space Ranger outputs into R and create `SpatialExperiment` objects.

Note that we create a separate `SpatialExperiment` object for each sample, and store these in a list.

```{r, message=FALSE}
library(SpatialExperiment)
library(Matrix)
library(rjson)
library(dplyr)
```

```{r}
sample_names <- c("LC_1", "LC_2")
spe_list <- vector("list", length = length(sample_names))
names(spe_list) <- sample_names

for (i in seq_along(sample_names)) {
  
  # ---------
  # load data
  # ---------
  
  sample_name <- sample_names[i]
  
  # path to Space Ranger output files (either on JHPCE or Mac laptop)
  if (Sys.info()["sysname"] == "Linux") {
    # files on JHPCE cluster
    dir_outputs <- "../outputs/NextSeqMiSeq"
  } else if (Sys.info()["sysname"] == "Darwin") {
    # files on Mac laptop
    dir_outputs <- "~/data/locus-c/outputs/NextSeqMiSeq"
  }
  
  # using "filtered" output files containing only spots over tissue
  dir_matrix <- file.path(dir_outputs, sample_name, "outs", "filtered_feature_bc_matrix")
  
  # barcodes
  file_barcodes <- file.path(dir_matrix, "barcodes.tsv.gz")
  df_barcodes <- read.csv(file_barcodes, sep = "\t", header = FALSE, 
                          col.names = c("barcode_id"))
  # features
  file_features <- file.path(dir_matrix, "features.tsv.gz")
  df_features <- read.csv(file_features, sep = "\t", header = FALSE, 
                          col.names = c("gene_id", "gene_name", "feature_type"))
  # counts
  file_counts <- file.path(dir_matrix, "matrix.mtx.gz")
  counts <- readMM(file = file_counts)
  
  # spatial coordinates
  dir_spatial <- file.path(dir_outputs, sample_name, "outs", "spatial")
  file_tisspos <- file.path(dir_spatial, "tissue_positions_list.csv")
  df_tisspos <- read.csv(file_tisspos, header = FALSE, 
                         col.names=c("barcode_id", "in_tissue", "array_row", "array_col", 
                                     "pxl_col_in_fullres", "pxl_row_in_fullres"))
  
  # image paths
  imagepaths <- file.path(dir_spatial, c("tissue_hires_image.png", "tissue_lores_image.png"))
  
  # spatial scale factors
  file_scale <- file.path(dir_spatial, "scalefactors_json.json")
  scalefactors <- fromJSON(file = file_scale)
  
  # check dimensions
  dim(df_barcodes)
  dim(df_features)
  dim(counts)
  # df_tisspos contains all spots (not filtered) - need to match rows
  dim(df_tisspos)
  
  
  # ------------------------
  # create SpatialExperiment
  # ------------------------
  
  # column data: match barcode IDs in 'df_barcodes' and 'df_tisspos'
  dim(df_barcodes)
  dim(df_tisspos)
  stopifnot(all(df_barcodes$barcode_id %in% df_tisspos$barcode_id))
  
  col_data <- left_join(df_barcodes, df_tisspos, by = "barcode_id")
  rownames(col_data) <- col_data$barcode_id
  dim(col_data)
  head(col_data, 3)
  
  # spatial data
  spatial_data <- col_data[, c("pxl_col_in_fullres", "pxl_row_in_fullres")]
  colnames(spatial_data) <- c("x", "y")
  head(spatial_data, 3)
  
  # include same column names in colData and spatialData for easier plotting
  col_data <- cbind(col_data, spatial_data)
  
  # row data
  row_data <- df_features
  rownames(row_data) <- df_features$gene_id
  head(row_data, 3)
  
  # note: not using image files or scale factors for now
  
  # create object
  spe <- SpatialExperiment(
    rowData = row_data, 
    colData = col_data, 
    spatialData = spatial_data, 
    assays = c(counts = counts), 
    sample_id = sample_names[i]
  )
  
  # store in list
  spe_list[[i]] <- spe
}

spe_list
```


## Quality control (QC) metrics

Calculate quality control (QC) metrics per sample using `scater`.

```{r, message=FALSE}
library(scater)
library(ggplot2)
```

```{r}
for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  # identify mitochondrial genes
  is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
  table(is_mito)
  rowData(spe)$gene_name[is_mito]
  
  # remove any spots with zero reads
  is_zero <- colSums(counts(spe)) == 0
  if (sum(is_zero) > 0) {
    spe <- spe[, !is_zero]
  }
  
  # calculate QC metrics using scater package
  spe <- addPerCellQC(spe, subsets = list(mito = is_mito))
  
  head(colData(spe), 3)
  
  # store in list
  spe_list[[i]] <- spe
}

# check
head(colData(spe_list[[1]]), 3)
sapply(spe_list, dim)
```


## Plot QC metrics

```{r, fig.width=4.75, fig.height=3.5}
# plot UMI counts
for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  p <- ggplot(as.data.frame(colData(spe)), 
              aes(x = y, y = x, color = sum)) + 
    geom_point(size = 0.5) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient(low = "gray90", high = "navy") + 
    ggtitle(paste0("UMI counts: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}

# plot number of detected genes
for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  p <- ggplot(as.data.frame(colData(spe)), 
              aes(x = y, y = x, color = detected)) + 
    geom_point(size = 0.5) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient(low = "gray90", high = "deeppink3") + 
    ggtitle(paste0("Detected genes: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}

# plot percent mitochondrial reads
for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  p <- ggplot(as.data.frame(colData(spe)), 
              aes(x = y, y = x, color = subsets_mito_percent)) + 
    geom_point(size = 0.5) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_gradient(low = "gray90", high = "darkorange2") + 
    ggtitle(paste0("Percent mitochondrial: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}

# spots with high mitochondrial reads
for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  thresh_mito <- 35
  colData(spe)$mito_high <- colData(spe)$subsets_mito_percent > thresh_mito
  
  p <- ggplot(as.data.frame(colData(spe)), 
              aes(x = y, y = x, color = mito_high)) + 
    geom_point(size = 0.5) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_manual(values = c("gray90", "red")) + 
    ggtitle(paste0("High mitochondrial percent: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```


Spots with high mitochondrial percentages are mostly within the white matter regions, so we leave them.

```{r}
for (i in seq_along(sample_names)) {
  spe <- spe_list[[i]]
  print(max(colData(spe)$subsets_mito_percent))
}
```


## Segment regions

Segment regions using threshold on number of UMI counts per spot. The region with high UMI counts is assumed to contain the LC neurons, and the region with low UMI counts is white matter.

```{r, fig.width=4.75, fig.height=3.5}
# select thresholds
thresh_umi <- c(3000, 1500)

for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  # identify spots above threshold
  colData(spe)$is_high_umi <- colData(spe)$sum > thresh_umi[i]
  spe_list[[i]] <- spe
  
  # plot thresholded regions
  p <- ggplot(as.data.frame(colData(spe)), 
              aes(x = y, y = x, color = is_high_umi)) + 
    geom_point(size = 0.5) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_manual(values = c("gray90", "red")) + 
    ggtitle(paste0("High UMI counts: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
}
```

```{r}
# lists to store SpatialExperiments containing segmented regions
spe_neuro_list <- spe_wm_list <- vector("list", length = length(sample_names))
names(spe_neuro_list) <- names(spe_wm_list) <- sample_names

for (i in seq_along(sample_names)) {
  
  spe <- spe_list[[i]]
  
  spe_neuro <- spe[, colData(spe)$is_high_umi]
  spe_wm <- spe[, !colData(spe)$is_high_umi]
  
  dim(spe_neuro)
  dim(spe_wm)
  
  spe_neuro_list[[i]] <- spe_neuro
  spe_wm_list[[i]] <- spe_wm
}
```


## Analyses within regions

Now continue with the analysis pipeline within the region containing the LC neurons identified above.


### Normalization

```{r, message=FALSE}
library(scran)
```

```{r, fig.width=4.5, fig.height=4}
for (i in seq_along(sample_names)) {
  
  # using subsetted object containing LC neurons region
  spe <- spe_neuro_list[[i]]
  dim(spe)
  
  # calculate pool-based size factors
  set.seed(123)
  qclus <- quickCluster(spe)
  table(qclus)
  spe <- computeSumFactors(spe, cluster = qclus)
  
  summary(sizeFactors(spe))
  hist(log10(sizeFactors(spe)), 
       xlab = "log10 (size factors)", main = "Size factors")
  
  # calculate logcounts
  spe <- logNormCounts(spe)
  
  # store in list
  spe_neuro_list[[i]] <- spe
}
```


### Feature selection

Feature selection using highly variable genes (HVGs).

```{r, fig.width=4.5, fig.height=4}
top_hvgs_list <- vector("list", length = length(sample_names))
names(top_hvgs_list) <- sample_names

for (i in seq_along(sample_names)) {
  
  # using subsetted object containing LC neurons region
  spe <- spe_neuro_list[[i]]
  dim(spe)
  
  # remove mitochondrial genes
  is_mito <- grepl("(^MT-)|(^mt-)", rowData(spe)$gene_name)
  table(is_mito)
  rowData(spe)$gene_name[is_mito]
  spe <- spe[!is_mito, ]
  dim(spe)
  
  # remove low-count genes
  is_low <- rowSums(counts(spe)) <= 5
  table(is_low)
  spe <- spe[!is_low, ]
  dim(spe)
  
  # store in list
  spe_neuro_list[[i]] <- spe
  
  # fit mean-variance relationship
  dec <- modelGeneVar(spe)
  fit <- metadata(dec)
  plot(fit$mean, fit$var, 
       xlab = "mean of log-expression", ylab = "variance of log-expression")
  curve(fit$trend(x), col = "dodgerblue", add = TRUE, lwd = 2)
  
  # select top HVGs
  top_hvgs <- getTopHVGs(dec, prop = 0.1)
  length(top_hvgs)
  
  # store in list
  top_hvgs_list[[i]] <- top_hvgs
}
```


### Dimensionality reduction

```{r}
for (i in seq_along(sample_names)) {
  
  # using subsetted object containing LC neurons region
  spe <- spe_neuro_list[[i]]
  dim(spe)
  
  top_hvgs <- top_hvgs_list[[i]]
  length(top_hvgs)
  
  # calculate PCA
  set.seed(123)
  spe <- runPCA(spe, subset_row = top_hvgs)
  
  reducedDimNames(spe)
  dim(reducedDim(spe, "PCA"))
  
  # calculate UMAP on top 50 PCs
  set.seed(123)
  spe <- runUMAP(spe, dimred = "PCA")
  # update column names
  colnames(reducedDim(spe, "UMAP")) <- paste0("UMAP", 1:2)
  
  reducedDimNames(spe)
  dim(reducedDim(spe, "UMAP"))
  
  # store in list
  spe_neuro_list[[i]] <- spe
}
```


### Clustering

```{r}
for (i in seq_along(sample_names)) {
  
  # using subsetted object containing LC neurons region
  spe <- spe_neuro_list[[i]]
  dim(spe)
  
  # graph-based clustering
  set.seed(123)
  g <- buildSNNGraph(spe, use.dimred = "PCA")
  g_walk <- igraph::cluster_walktrap(g)
  clus <- g_walk$membership
  table(clus)
  
  colLabels(spe) <- factor(clus)
  
  # store in list
  spe_neuro_list[[i]] <- spe
}
```


```{r, fig.width=5, fig.height=4.5}
for (i in seq_along(sample_names)) {
  
  # using subsetted object containing LC neurons region
  spe <- spe_neuro_list[[i]]
  dim(spe)
  
  # color palette
  colors <- unname(palette.colors(palette = "Okabe-Ito"))
  
  
  # plot clusters in spatial coordinates
  p <- ggplot(as.data.frame(colData(spe)), 
              aes(x = y, y = x, color = label)) + 
    geom_point(size = 1.0) + 
    coord_fixed() + 
    scale_y_reverse() + 
    scale_color_manual(values = colors) + 
    ggtitle(paste0("Clustering: spatial coordinates: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
  
  fn <- file.path("../../plots", paste0("clustering_segmented_XY_", sample_names[i], ".png"))
  ggsave(fn, width = 5, height = 4.5)
  
  
  # plot clusters in UMAP coordinates
  p <- ggplot(as.data.frame(cbind(colData(spe), reducedDim(spe, "UMAP"))), 
              aes(x = UMAP1, y = UMAP2, color = label)) + 
    geom_point(size = 1.0) + 
    scale_y_reverse() + 
    scale_color_manual(values = colors) + 
    ggtitle(paste0("Clustering: UMAP coordinates: ", sample_names[i])) + 
    theme_bw() + 
    theme(panel.grid = element_blank(), 
          axis.title = element_blank(), 
          axis.text = element_blank(), 
          axis.ticks = element_blank())
  
  print(p)
  
  fn <- file.path("../../plots", paste0("clustering_segmented_UMAP_", sample_names[i], ".png"))
  ggsave(fn, width = 5, height = 4.5)
}
```


### Marker genes

Identify top marker genes for clusters identified above.

```{r, message=FALSE}
library(pheatmap)
```

```{r, fig.width=4, fig.height=4}
top_genes_list <- vector("list", length = length(sample_names))
names(top_genes_list) <- sample_names

for (i in seq_along(sample_names)) {
  
  print(paste0("Sample ", sample_names[i]))
  
  # using subsetted object containing LC neurons region
  spe <- spe_neuro_list[[i]]
  # set gene names as row names for easier visualization in plots
  rownames(spe) <- rowData(spe)$gene_name
  
  # test for marker genes
  markers <- findMarkers(spe, test = "binom", direction = "up")
  # list with one DataFrame per cluster
  markers
  
  top_genes_list[[i]] <- vector("list", length = length(markers))
  names(top_genes_list[[i]]) <- names(markers)
  
  for (k in seq_along(markers)) {
    
    # plot log-fold changes for one cluster over all other clusters
    interesting <- markers[[k]]
    best_set <- interesting[interesting$Top <= 5, ]
    logFCs <- getMarkerEffects(best_set)
    
    pheatmap(logFCs, cluster_cols = FALSE, 
             breaks = seq(-5, 5, length.out = 101), 
             main = paste0(sample_names[i], ": ", "cluster ", k))
    
    # plot log-transformed normalized expression of top genes for one cluster
    top_genes <- rownames(best_set)
    
    top_genes_list[[i]][[k]] <- top_genes
  }
}
```

```{r}
p_list <- vector("list", length = length(sample_names))
names(p_list) <- sample_names

for (i in seq_along(sample_names)) {
  
  print(paste0("Sample ", sample_names[i]))
  
  # using subsetted object containing LC neurons region
  spe <- spe_neuro_list[[i]]
  # set gene names as row names for easier visualization in plots
  rownames(spe) <- rowData(spe)$gene_name
  
  # get combined set of top genes for all clusters
  top_genes <- unique(unlist(top_genes_list[[i]]))
  # sort alphabetically
  top_genes <- sort(top_genes)
  length(top_genes)
  
  # plot expression of combined set of top genes
  p <- plotExpression(spe, x = "label", features = top_genes) + 
    ggtitle(sample_names[i])
  
  p_list[[i]] <- p
}
```

```{r, fig.width=7.5, fig.height=50}
# display plots using custom width and height
print(p_list[[1]])
```

```{r, fig.width=5, fig.height=14}
# display plots using custom width and height
print(p_list[[2]])
```


## Spatial expression plots

### Top HVGs

Plot spatial expression for top (non-spatial) highly variable genes (HVGs) identified in analysis pipeline above.

```{r, fig.width=3.5, fig.height=3.25}
for (i in seq_along(sample_names)) {
  
  print(paste0("Sample ", sample_names[i]))
  
  # using subsetted object containing LC neurons region
  spe <- spe_neuro_list[[i]]
  
  # get top n HVGs
  n <- 100
  top_hvgs <- top_hvgs_list[[i]][1:n]
  names(top_hvgs) <- rowData(spe)[top_hvgs, ][, "gene_name"]
  print(names(top_hvgs))
  # sort alphabetically
  ord <- order(names(top_hvgs))
  top_hvgs <- top_hvgs[ord]
  
  # subset object
  spe_sub <- spe[top_hvgs, ]
  dim(spe_sub)
  
  max_logcounts <- max(logcounts(spe_sub))
  
  for (k in seq_along(top_hvgs)) {
    
    # plot spatial distribution of expression per gene
    df <- cbind(as.data.frame(colData(spe)), 
                expr = counts(spe)[top_hvgs[k], ], 
                logexpr = logcounts(spe)[top_hvgs[k], ])
    
    p <- ggplot(df, aes(x = y, y = x, color = logexpr)) + 
      geom_point(size = 0.5) + 
      coord_fixed() + 
      scale_y_reverse() + 
      scale_color_gradient(limits = c(0, max_logcounts), low = "gray90", high = "blue3") + 
      ggtitle(paste0(sample_names[i], ": ", names(top_hvgs)[k])) + 
      theme_bw() + 
      theme(panel.grid = element_blank(), 
            axis.title = element_blank(), 
            axis.text = element_blank(), 
            axis.ticks = element_blank())
    
    print(p)
  }
}
```


### Cluster marker genes

Plot spatial expression for cluster marker genes identified in analysis pipeline above.

```{r, fig.width=3.5, fig.height=3.25}
for (i in seq_along(sample_names)) {
  
  print(paste0("Sample ", sample_names[i]))
  
  # using subsetted object containing LC neurons region
  spe <- spe_neuro_list[[i]]
  # set gene names as row names for easier visualization in plots
  rownames(spe) <- rowData(spe)$gene_name
  
  # get combined set of top genes for all clusters
  top_genes <- unique(unlist(top_genes_list[[i]]))
  # sort alphabetically
  top_genes <- sort(top_genes)
  length(top_genes)
  
  print(top_genes)
  
  # subset object
  spe_sub <- spe[rowData(spe)$gene_name %in% top_genes, ]
  dim(spe_sub)
  
  max_logcounts <- max(logcounts(spe_sub))
  
  for (k in seq_along(top_genes)) {
    
    # plot spatial distribution of expression per gene
    df <- cbind(as.data.frame(colData(spe)), 
                expr = counts(spe)[top_genes[k], ], 
                logexpr = logcounts(spe)[top_genes[k], ])
    
    p <- ggplot(df, aes(x = y, y = x, color = logexpr)) + 
      geom_point(size = 0.5) + 
      coord_fixed() + 
      scale_y_reverse() + 
      scale_color_gradient(limits = c(0, max_logcounts), low = "gray90", high = "red3") + 
      ggtitle(paste0(sample_names[i], ": ", top_genes[k])) + 
      theme_bw() + 
      theme(panel.grid = element_blank(), 
            axis.title = element_blank(), 
            axis.text = element_blank(), 
            axis.ticks = element_blank())
    
    print(p)
  }
}
```


### Calculate SVGs

Calculate spatially variable genes (SVGs) using SPARK-X (Sun et al. 2020: https://xzhoulab.github.io/SPARK/).

Notes:
- SPARK-X is faster version of SPARK, which scales to Visium-sized datasets
- SPARK-X (and SPARK) use raw counts (not logcounts)

```{r, message=FALSE}
library(SPARK)
```

```{r}
for (i in seq_along(sample_names)) {
  
  print(paste0("Sample ", sample_names[i]))
  
  # using subsetted object containing LC neurons region
  spe <- spe_neuro_list[[i]]
  
  # inputs for SPARK-X
  # note mitochondrial genes have been filtered out (as recommended)
  sp_count <- counts(spe)
  location <- spatialCoords(spe)
  dim(sp_count)
  dim(location)
  
  # run SPARK
  sparkX <- sparkx(sp_count, location)
  
  # store outputs
  stopifnot(all(rownames(sparkX$res_mtest) == rownames(spe)))
  rowData(spe) <- cbind(rowData(spe), DataFrame(sparkX$res_mtest))
  
  spe_neuro_list[[i]] <- spe
}
```


### Plot SVGs

Plot spatial expression patterns for spatially variable genes (SVGs) from SPARK-X calculated above.

```{r, fig.width=3.5, fig.height=3.25}
for (i in seq_along(sample_names)) {
  
  print(paste0("Sample ", sample_names[i]))
  
  # using subsetted object containing LC neurons region
  spe <- spe_neuro_list[[i]]
  # set gene names as row names for easier visualization in plots
  rownames(spe) <- rowData(spe)$gene_name
  
  # subset object containing top n SVGs
  n <- 100
  ix_top <- which(rank(rowData(spe)$adjustedPval, ties.method = "first") <= n)
  spe_sub <- spe[ix_top, ]
  dim(spe_sub)
  
  # names of top n SVGs in ranked order
  ord <- order(rowData(spe_sub)$adjustedPval)
  top_svgs <- rowData(spe_sub)$gene_name[ord]
  print(top_svgs)
  
  # alphabetical order for plotting
  top_svgs_alph <- sort(top_svgs)
  
  max_logcounts <- max(logcounts(spe_sub))
  
  for (k in seq_along(top_svgs_alph)) {
    
    # plot spatial distribution of expression per gene
    df <- cbind(as.data.frame(colData(spe_sub)), 
                expr = counts(spe_sub)[top_svgs_alph[k], ], 
                logexpr = logcounts(spe)[top_svgs_alph[k], ])
    
    p <- ggplot(df, aes(x = y, y = x, color = logexpr)) + 
      geom_point(size = 0.5) + 
      coord_fixed() + 
      scale_y_reverse() + 
      scale_color_gradient(limits = c(0, max_logcounts), low = "gray90", high = "green4") + 
      ggtitle(paste0(sample_names[i], ": ", names(top_svgs_alph)[k])) + 
      theme_bw() + 
      theme(panel.grid = element_blank(), 
            axis.title = element_blank(), 
            axis.text = element_blank(), 
            axis.ticks = element_blank())
    
    print(p)
  }
}
```


# Session info

```{r, message=FALSE}
sessionInfo()
```

